---
layout: page
title: Data Types and Structures
order: 4
session: 1
length: 20
toc: true
adapted: true
attrib_name: Programming with R
attrib_link: https://github.com/swcarpentry/r-novice-inflammation
attrib_copywrite: Software Carpentry
attrib_license: CC-BY 4.0
attrib_license_link: https://creativecommons.org/licenses/by/4.0/
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


R has a number of data sets included for you to practice with. 
We can get a list of these by running the command ```data()```

```{r echo=TRUE}
data()
```

# Exploring datasets

In this session, we will be using the built-in R data sets stored in the variables “iris” & "mtcars". 
By having our data stored in a variable we can manipulate it or start doing this with it. By this we mean applying functions to it. Note that we are using the word data, very loosely, it can refer to any form of information we want to store.

Let's take a closer look at the `iris` data.

First, let's ask what type of thing `iris` is using the function ```class()```

```{r}
class(iris)
```

The output tells us that it is a data frame. Think of this structure as a spreadsheet in MS Excel that many of us are familiar with.
Data frames are very useful for storing data and you will find them elsewhere when programming in R. A typical data frame of
 experimental data contains individual observations in rows and variables in columns.

We can see the shape, or dimensions, of the data frame with the function ```dim()```:

```{r}
dim(iris)
```

This tells us that our data frame, `iris`, has `r nrow(iris)` rows and `r ncol(iris)` columns.


To explore data frames, there are a number of relevant functions:

* ```head()``` - shows first 6 rows
* ```tail()``` - shows last 6 rows
* ```dim()``` - returns the dimensions of data frame (i.e. number of rows and number of columns)
* ```nrow()``` - number of rows
* ```ncol()```- number of columns
* ```str()``` - structure of data frame - name, type and preview of data in each column
* ```names()``` or ```colnames()``` - both show the names attribute for a data frame


## Activity: Data Exploration

Use these functions to explore the mtcars dataset



## Subsetting Data

If we want to get a single value from the data frame, we can provide an index in square brackets, just as we do in math. This is called slicing.

```{r}
# first value in dat
iris[1, 1]
# middle value in dat
iris[30, 20]
```

An index like ``[30, 20]``` selects a single element of a data frame, but we can select larger sections as well.
For example, we can select the first three columns of values for the first four rows like this:

```{r}
iris[1:4, 1:3]
```

The slice ```1:4``` means, "Start at index 1 and go to index 4."

The slice does not need to start at 1, e.g. the line below selects rows 5 through 10:

```{r}
iris[5:10, 1:3]
```
We can use the function ```c()```, which stands for **c**ombine, to select non-contiguous values:

```{r}
iris[c(3, 8, 37, 56), c(1,3)]
```

We also don't have to provide a slice for either the rows or the columns.
If we don't include a slice for the rows, R returns all the rows; if we don't include a slice for the columns, R returns all the columns.
If we don't provide a slice for either rows or columns, e.g. `iris[, ]`, R returns the full data frame.

```{r}
# All columns from row 5
iris[5, ]
# All rows from column 2
iris[, 2]
```

<details>
	<summary> Addressing Columns by Name </summary>
	<pre>

    Columns can also be addressed by name, with either the `$` operator (ie. `iris$Petal.Length`) or square brackets (ie. `iris[,'Petal.Length']`).
    </pre>
</details>





# Data Types 

*Data types*, or modes, define what the values are and how they can be used. 
There are __*5 core data types*:__

* Numeric - all real numbers e.g. 7.5
* Integer - e.g 2
* Complex - numbers with real and imaginary parts e.g 1+4i
* Character - consists of letters or numbers or symbols or a combination of these e.g. "a", "f5", "datatypes", "Learning R is fun".
* Logical - only takes TRUE or FALSE

All data will we associated to a specific type. R provides many functions to examine features of objects. Some are listed here:

* ```typeof()``` - what is the object’s data type (on the data storage level ("what the computer sees"))?
* ```class()``` - what is the object's data type (on the abstract type level("what R sees"))?
* ```mode()``` - what is the object's data type (on the data storage level ("what the computer sees"))?
* ```length()```- how long is it? 
* ```attributes()``` - does it have any metadata? 
* ```str``` - display the internal structure of an object.
* ```is.numeric()```, ```is.character()```, ```is.complex()```, ```is.logical()``` - returns TRUE when an object is the datatype queried, FALSE if not

__Some Examples:__
```{r, echo=TRUE, eval=TRUE, error=TRUE}
x <- "dataset"
typeof(x)
```

```{r, echo=TRUE, eval=TRUE, error=TRUE}
attributes(x)
```

```{r, echo=TRUE, eval=TRUE, error=TRUE}
y <- 1:10
y
```
```{r, echo=TRUE, eval=TRUE, error=TRUE}
typeof(y)
```

```{r, echo=TRUE, eval=TRUE, error=TRUE}
length(y)
```

#### Task 4 & 5 <br>{-}
>4.) Create the following objects:<br>```x=8```<br>``` y=8.9```<br> Use both ```typeof()``` and ```mode()``` to determine the datatypes of ```x``` and ```y```. What differences do you observe? \
5.) When you use ```class()``` on the ```iris``` dataset (*e.g.* ```class(iris)```) and when you use ```typeof()``` on the iris data set, which results do you get? The resulting data types/structures will be explained later in detail; keep the results in mind for [4.1.4.3](#4.1.4.3 Lists) and [4.1.4.4.](#4.1.4.4) <br>

## Converting between datatypes

It can be critical that R has correctly assigned the right data type to your variable. If it has not you may run into errors when processing it. You therefore may want to convert between different data types. This can be done with the series of functions `as.numeric`, `as.character` etc. 

For example, let's convert our numeric variable `y` into a character. 

```{r, echo=TRUE, eval=TRUE, error=TRUE}
y <- as.character(y)
typeof(y)
```

# Data Structures

R has a number of inbuilt strutures that can be used to store datasets. We have encountered one of these already the data.frame. Other include: 

* strings
* vectors 
* data frames
* matrices
* arrays
* lists
* factors

### strings

A string is a run of characters. e.g. "hello". or "a189jde2mjo". They are enclosed in quotes.

### vectors

A vector is the most common and basic data structure in R. They are an ordered collection of basic data types of a given length. They are one-dimensional.

The concatenate or combine ```c()``` function will explicitly __construct__ a vector. 

```{r, echo=TRUE, eval=TRUE, error=TRUE}
v_num <- c(9,20,12)
v_log <- c(TRUE, FALSE, FALSE, TRUE)
```

As we create these vectors you should see them listed in your environment pane. We can then call the name of the varible to see what the value is at that point in time. 

```{r, echo=TRUE, eval=TRUE, error=TRUE}
v_num 
v_log 
```

The function ```c()``` can also be used to __add elements__ to a vector.


```{r, echo=TRUE, eval=TRUE, error=TRUE}
v_name <- c("Sarah", "Tracy", "Jon")
v_name <- c(v_name, "Annette")
v_name
```

If we want to create a series of numbers, like 1 to 140, we can bypass the `c()` constructor and just write


```{r, echo=TRUE, eval=TRUE, error=TRUE}
(v_int <- 1:4)
```

When we call a vector *atomic*, we mean that the vector only holds data of a single data type.

### data.frames

A data frame is a two dimensional structure consisting of rows and columns. 

We can create a data frames using the function `data.frame()`. We will use the predefined vector `letter` to get the first ten letters in the alphabet.

```{r, echo=TRUE, eval=TRUE, error=TRUE}
dat <- data.frame(id = letters[1:10], x = 1:10, y = 11:20)
dat
```

Columns are variables, and each column will have a specified data type, which all entries must be. Rows are observations. Rows and columns will have `rownames` and `colnames` that you can use to extract specific rows or columns respectively. 

If you read in a table of data from a file, it will typically be represented by a data.frame.

##### Functions to explore data frames with:{-}

* ```head()``` - shows first 6 rows
* ```tail()``` - shows last 6 rows
* ```dim()``` - returns the dimensions of data frame (i.e. number of rows and number of columns)
* ```nrow()``` - number of rows
* ```ncol()```- number of columns
* ```str()``` - structure of data frame - name, type and preview of data in each column
* ```names()``` or ```colnames()``` - both show the names attribute for a data frame

> ## Task
> 
> Use the functions above on the objects `dat` and `mtcars`.

### matrices

A matrix is another two dimensional object, but it differs to a data.frame as all column/entries must be of the same type. It is more efficient memory wise than a data.frame.

We can construct a matrix as follows:

```{r, echo=TRUE, eval=TRUE, error=TRUE}
m<-matrix(1:6, nrow=2, ncol=3)
m
```

When creating a __matrix__, it is important to remember that matrices __*are filled column-wise*__ If that is not what you want, you can use the ```byrow``` argument (a logical: can be ```TRUE``` or ```FALSE```) to specify how the matrix is filled

```{r, echo=TRUE, eval=TRUE, error=TRUE}
m<-matrix(1:6, nrow=2, ncol=3, byrow=TRUE)
m
```

We can confirm the data type of a matrix with the function `type`.


```{r, echo=TRUE, eval=TRUE, error=TRUE}
class(m)
```

### arrays

Arrays are n dimensional storage structures. A one dimensional array is a vector, a two dimensional array is a matrix. 

We will not be using this type of object, but included for completeness.

### lists

A list in R is a collection of objects and elements, which themselves can be a heterogeneous mix of other objects including vectors, matrixes, data.frames, functions, strings, numbers. They tend to be used to collate different data types connected in some way. 

We can create a list of the vectors, data.frames we have already constructed as follows.


```{r, echo=TRUE, eval=TRUE, error=TRUE}
l <- list(v_log, v_num, v_it, iris)
```

We will not be using this type of object, but included here for completeness.

### factors

Sometimes considered as a data type, as confusing a possible response to the function `type`, and a possible option for columns in a data frame. A factor is how categorical variables are stored. They have a finite number of "levels", which are the options that any elements of that data type can take. They are actually stored as integers which can make them quite powerful for subsetting. 

It is very easy for character vectors to be inappropriately stored as factors and vice versa. 

For example, if we define a vector of months, the default is to consider a vector of characters. We have to actively coerce it into a factor. 

```{r, echo=TRUE, eval=TRUE, error=TRUE}
a      <- c("March","February","February","November","February","March","March","March","February","November")
class(a)
fact <- as.factor(a)
class(a)
```


#### Rearranging level order{-}
\
__*A factor will define a hierarchy for its levels*__. The default here is to order alphabetically. This order is the basic for any tables or plots ( e.g. boxplot) we generate with this variable. If we want these outputs to in a __different order__ (i.e. N first followed by F then M) we would need to change the order of the levels as follows:

```{r, echo=TRUE, eval=TRUE, error=TRUE}
fact <- factor(a, levels=c("November","February","March"))
str(fact)
```
-------